\MakeUppercase{Linear versus centred Colouring Numbers}\thanks{This research was partly funded by NSERC.}

### 2022-04-04 lin-vs-cen.tex
Pfffft

The \emph{centred chromatic number} $\chicen(G)$ of $G$ is the minimum integer $c$ such that $G$ has a centred $c$-colouring.  We note that an easy argument shows that centered chromatic number and treedepth are equivalent, i.e., $\td(G)=\chicen(G)$ for every graph $G$. The \emph{linear chromatic number} of $G$ is the minimum integer $\ell$ such that $G$ has a linear $\ell$-colouring.  Since every path in $G$ is a connected subgraph of $G$, every centred colouring of $G$ is also a linear colouring of $G$, so $\chilin(G)\le\chicen(G)$.

@@

\begin{thm}[\citet{kun.obrien.ea:polynomial}]\label{kun_obrien_general}
  For any graph $G$, $\chicen(G)\le \chilin^{190}(G)\cdot\log^{O(1)}(\chilin(G))$.\todo{Update with most recent numbers}
\end{thm}

Their proof of \cref{kun_obrien_general} has three steps:
\begin{enumerate}
  \item A theorem of \citet{kawarabayashi.rossman:polynomial} shows that, if $\chicen(G)\ge k^{190}\log^{O(1)} k$ then $\chilin(G)\ge k^{38}$ or $G$ has treewidth $\tw(G)\ge k^{38}$.  In the former case there is nothing left to prove.
  \item The current-best version of the Excluded Grid Theorem due to \citet{chuzhoy:improved} shows that, if $\tw(G)\ge k^{38}$, then $G$ contains an $\Omega(k^2\times k^2)$ grid minor.
  \item A technical lemma \cite[Lemma~5]{kun.obrien.ea:polynomial} shows that, if $G$ contains a $k^2\times k^2$ grid minor, then $\chilin(G)\in\Omega(k)$.
\end{enumerate}

These notes are an attempt to improve the bound in \cref{kun_obrien_general} in the general case as well as for some special classes of graphs. First we point out that some new results can already reduce the exponent from $190$ to $39$ and eliminate the polylog factor.  (These improvements are already described by \citet[Theorem~1.9]{czerwinski.nadara.ea:improved}.)

First, \citet{chuzhoy.tan:towards} improved the exponent in the Excluded Grid Theorem:  There exists a constant $c$ such that any graph $G$ with $\tw(G)\ge k^{18}\log^c k$ has a $k\times k$ grid minor.  This already improves the bound from $k^{190}$ to $k^{90}$.
Next, a recent result of \citet[Theorem~1.4]{czerwinski.nadara.ea:improved} shows that if $b$ is the maximum treedepth of any tree contained in $G$ then $\td(G)\le b\cdot\tw(G)$ (and the proof is easy).  Therefore, if $\td(G)\ge k^{19}$ then at least one of the following cases applies:
\begin{enumerate}
    \item $\tw(G)\ge k^{18}$: In this case the proof follows Steps~2 and 3 as before, but using the grid minor result of \citet{chuzhoy.tan:towards} to show that $\chilin(G)\in\Omega(k)$.
    \item $G$ contains a tree $T$ of treedepth $k$.  The same paper \cite[Lemma~1.6]{czerwinski.nadara.ea:improved} shows that any tree $T$ of treedepth $k$ contains a subcubic subtree of treedepth $\Omega(k)$.
    A theorem of \citet{kun.obrien.ea:polynomial} shows that, for any subcubic tree $T$, $\chicen(T) \le (\log_2 3)\cdot\chilin(T)$.  Therefore, in this case $\chilin(G)\ge\chilin(T)\ge \chicen(T)/\log_2 3 = \Omega(k)$.
\end{enumerate}

This gives the following improvement to \cref{kun_obrien_general}
\begin{thm}[\citet{czerwinski.nadara.ea:improved}]\label{kun_obrien_general2}
  For any graph $G$, $\chicen(G)\in O(\chilin^{19}(G))$.
\end{thm}

As a next step, we would like to improve the third part of the argument to show that, if $G$ contains a $k\times k$ grid minor, then $\chilin(G)\in \Omega(k)$.  This, by itself, would reduce the exponent in \cref{kun_obrien_general2} from $39$ to $20$.  We observe that the first two parts of the argument are both ``heavy hammers'' and that, for some special graph classes, much lighter hammers suffice.  For example, if we consider only planar graphs, then it is well known that any planar graph $G$ of treewidth $k$ contains an $\Omega(k\times k)$ grid minor.  This already reduces the exponent in \cref{kun_obrien_general2} further (for planar graphs) to $3$ ($2$ if we can prove the result for grid minors).  If one could prove a similar result for the first ``heavy hammer'' then this would show that, for any planar graph $G$, $\chicen(G)\in O(\chilin(G))$.

@@

\begin{lem}\label{weighted_lovasz}
  Let $E:=\{E_1,\ldots,E_n\}$ be a set of events in some probability space $(\Omega,\Pr)$.  For each $i\in\{1,\ldots,n\}$, let $\Gamma_i\subseteq E$ be such that the event $E_i$ is mutually independent of $\Gamma_i$,\footnote{An event $A$ is mutually independent of a set $\{B_1,\ldots,B_r\}$ of events if, for any disjoint sets $I,J\subseteq\{1,\ldots,r\}$, $\Pr(A\cap\bigcap_{i\in I} B_i\cap\bigcap_{j\in J} \overline{B}_j=\Pr(A)\Pr(\bigcap_{i\in I} B_i\cap\bigcap_{j\in J} \overline{B}_j)$.} and let $w:E\to(0,1)$ be such that
  \[
      \Pr(E_i) \le w(E_i)\cdot\prod_{E_j\in\Gamma_i}(1-w(E_j))  \enspace .
  \]
  Then $\Pr(\overline{E_1}\cap\cdots\cap\overline{E_n}) > 0$.
\end{lem}

@@

In this work, we need to work with pseudogrids.  An \defin{$a\times b$-pseudogrid} is any graph that can be obtained from $G_{a\times b}$ in the following way:
\begin{compactenum}
  \item Replace each edge $vw$ of $G_{a\times b}$ with a path $\overline{P}_{vw}$ whose endpoints are $v$ and $w$.  (In other words, $\overline{P}_{vw}$ is a path obtained by subdividing the edge $vw$ zero or more times.)
  \item Replace each degree-$4$ vertex $(x,y)$ of $G_{a\times b}$ with a path $P_v$. If $P_v$ has only one vertex then this operation has no effect.  Otherwise, $P_v$ has two endpoints $p$ and $q$ and
  \begin{compactenum}[(Q1)]
    \item \label{q_i} $p$ is adjacent to $\{(x-1,y), (x,y-1)\}$ and $q$ is adjacent to $(x+1,y),(x,y+1)$;
    \item \label{q_ii} $p$ is adjacent to $\{(x,y+1), (x-1,y)\}$ and $q$ is adjacent to $(x,y-1),(x+1,y)$; or
    \item \label{q_iii} $p$ is adjacent to $\{(x-1,y),(x+1,y)\}$ and $q$ is adjacent to $\{(x,y-1),(x,y+1)\}$.
  \end{compactenum}
\end{compactenum}

Let $G$ be an $a\times b$ pseudogrid.  For an edge $vw$ of $G_{a\times b}$, we let $P_{vw}:=\overline{P}_{vw}-\{v,w\}$ denote the (possibly empty) subpath containing the internal vertices of $\overline{P}_{vw}$.  For each vertex $v$ of $G_{a\times b}$ of degree less than $4$ we define $P_{v}$ to be the $1$-vertex path that contains only $v$.  In this way, $\mathcal{P}:=\{V(P_\mu):\mu\in \VE(G_{a\times b})\}$ is a partition of $V(G)$ into induced paths.  We call $\mathcal{P}$ the \defin{grid-partition} of $G$.
% and we denote it by $(P_\mu:\mu\in \VE(G_{a\times b}))$.
% This defines a mapping $\delta:V(G)\to V(G_{a\times b})\cup E(G_{a\times b})$ where $u\in V(P_{\delta(u)})$ for each $u\in v(G)$.

@@

Each row $R':=v_1,\ldots,v_a$ of $G_{a\times b}$ corresponds naturally to a path $R$ of $G$. The path $R$ contains $V(\overline{P}_{v_iv_{i+1}})$ for each $i\in\{1,\ldots,a-1\}$.  However, for $i\in\{2,\ldots,a-1\}$ $R$ may or may not contain $V(P_{v_i})$.  In particular, if $P_{v_i}$ was created using (Q\ref{q_iii})\todo{Fix this!} then $R$ does not contain any internal vertices in $P_{v_i}$.
Similarly, a column $C':=v_1,\ldots,v_b$ of $G_{a\times b}$ corresponds to a path that contains $V(\overline{P}_{v_jv_{j+1}})$ for each $j\in\{1,\ldots,b-1\}$. This correspondence allows us to talk about the rows and columns of $G$, which we will do next.

As part of our proof, we use the operation of \defin{deleting} a row (or column) $R$ of $G$, which works as follows:  We remove all edges of $R$ from $G$ and eliminate any isolated vertices.  If this produces vertices of degree $1$ (which happens when $R$ is the first or last row of $G$ or when $R=v_1,\ldots,v_a$ does not contains $P_{v_i}$ for some $i\in\{1,\ldots,r\}$) then we repeatedly remove degree-$1$ vertices until none remain.  If $G$ is an $a\times b$ pseudogrid and we delete some row $R$, then the resulting graph is an $a\times (b-1)$ pseudogrid.  Similarly, if we delete column $C$ of $G$, then the resulting graph is an $(a-1)\times b$ pseudogrid.

@@

For convenience, let $G_k:=G_{k\times k}$.
Let $G$ be a $k\times k$ pseudogrid with grid partition $\mathcal{P}:=(P_\mu:\mu\in\VE(G_{k}))$ and let $\varphi:V(G)\to\{1,\ldots,c\}$ be a vertex colouring of $G$.  For any $\mu\in\VE(G_{k})$, we let $\varphi_{\mathcal{P}}(\mu):=\{\varphi(v):v\in V(P_\mu)\}$.  For any colour $\alpha\in\{1,\ldots,c\}$, define $\varphi^{-1}(\alpha):=\{v\in V(G):\varphi(v)=\alpha\}$ and define $\varphi_\mathcal{P}^{-1}(\alpha):=\{\mu\in \VE(G_{k\times k}):\alpha\in\varphi_\mathcal{P}(\mu)\}$.  For any colour set $A\subseteq\{1,\ldots,c\}$ define $\varphi^{-1}(A):=\bigcup_{\alpha\in A}\varphi^{-1}(\alpha)$ and $\varphi_\mathcal{P}^{-1}(A):=\bigcup_{\alpha\in A}\varphi_{\mathcal{P}}^{-1}(\alpha)$.

@@

\begin{proof}
  The proof is by induction on $|\varphi(V(G))|$, the number of colours used by the colouring $\varphi$.  If there exists no $A\subseteq \varphi(V(G))$ with $\varphi_{\mathcal{P}}^{-1}(A) < d|A|$ then taking $G':=G$ and $\mathcal{P}':=\mathcal{P}$ satisfies the requirements of the lemma. Otherwise, we will remove a set $R$ of rows and a set $C$ of columns from $G$ with $|R|=|C|\le d|A|$ to eliminate all vertices with colours in $A$, as follows:
  \begin{compactitem}
    \item For each $v:=(i,j)\in V(G_{k})$ with $A\cap\varphi(v)\neq\emptyset$, we include row $j$ in $R$ and column $i$ in $C$.
    \item For each edge $vw\in E(G_{k\times k})$ with $A\cap\phi()\alpha\in\Phi(vw)$, we include the row of $G$ that contains $P_{vw}$ in $R$ or we include the column of $G$ that contains $P_{vw}$ in $C$.
    \item Finally, we add arbitrary rows to $R$ or columns to $C$ to ensure that $|R|=|C|$.
  \end{compactitem}
  Now we remove all rows in $R$ and all columns in $C$ from $G$ to obtain a $k_0\times k_0$ pseudogrid $G_0$ with $k_0\ge k-d|A|$ and such that $\varphi(V(G_0))\cap A=\emptyset$.  In particular, $|\varphi(V(G_0))|\le |\varphi(V(G))|-|A|$.

  Now apply induction on $G_0$ to get a $k'\times k'$ pseudogrid with
  \[  k'\ge k_0-d|\varphi(V(G_0))|  \ge k-d|A|-d|\varphi(V(G_0))|= k - d|\varphi(V(G))|
  \]
  that satisfies the conditions of the lemma.
\end{proof}

\begin{lem}\label{one_colour_per_object}
  Let $d>1$ be an integer, let $G$ be a $k\times k$ pseudogrid with grid-partition $\mathcal{P}$, and let $\varphi:V(G)\to\{1,\ldots,c\}$ be a vertex colouring of $G$ such that, for any $A\subseteq\varphi(V(G))$, $|\varphi_{\mathcal{P}'}^{-1}(A)| \ge d\cdot|A|$. Then there exists a (partial) colouring $\phi:\VE(G_{k})\to\{1,\ldots,c,\perp\}$ with the following properties:
  \begin{compactenum}[(i)]
    \item For each $\mu\in\VE(G_{k})$, $\phi(\mu)=\perp$ or $\phi(\mu)\in\varphi(\mu)$.
    \item For each $\alpha\in\varphi(V(G))$, $|\{\mu\in\VE(G_{k}):\phi(\mu)=\alpha\}|\ge d$.
  \end{compactenum}
\end{lem}

\begin{proof}
  Consider the bipartite graph $H$ with parts $X:=\{1,\ldots,c\}$ and $Y:=\VE(G_k)$ and edge set
  \[
    \{ (\alpha,\mu)\in X\times Y: \alpha\in\varphi(\mu) \}
  \]
  By \cref{d_hall}, $H$ contains a subgraph $M$ with $\deg_M(x)=d$ for each $\alpha\in X$ and $\deg_M(y)\le 1$ for each $\mu\in Y$.  For each edge $\alpha\mu$ in $M$, set $\phi(\mu):=\alpha$.  This defines $\phi(\mu)$ for any $\mu\in X$ with $\deg_M(\mu)=1$.  For each $\mu\in X$ with $\deg_M(\mu)=0$ set $\phi(\mu):=\perp$.
\end{proof}

@@

The \defin{$r$-box} centered at a vertex $v:=(i,j)$ of $G_{a\times b}$ is defined as
\[
  B_r(v) := \{i-r,\ldots,i+r\}\times\{j-r,\ldots,j+r\} \cap V(G_{a,b}) \enspace .
\]
The \defin{$r$-box} centered at an edge $vw$ of $G_{a\times b}$ is $B_r(vw):=B_r(v)\cup B_r(w)$.  Observe that, for any $\mu\in V(G_{a\times b})\cup E(G_{a\times b})$,  $v\in B_r(u)$ if and only if $u\in B_r(v)$.  Any $r$-box $B_r(\mu)$ defines an induced subgraph that we denote by $G_r(\mu):=G_{a\times b}[B_r(\mu)]$.  We extend these definitions to vertices of an $a\times b$ pseudogrid $G$ with grid-partition $\mathcal{P}:=\{V(P_\mu):\mu\in\VE(G)\}$. For any $v\in V(G)$, define
\[
   \tilde{B}_r(v) := \bigcup_{\mu\in \VE(G_r(\mu_v))} V(P_\mu)
\]
where $\mu_v$ denotes the unique element in $\VE(G_k)$ such that $v\in P_{\mu_v}$.
For any $v\in V(G)$, define $\tilde{G}_r(v):=G[\tilde{B}_r(v)]$.

@@

%   \begin{center}

@@

%   \end{center}

>>>>

The \emph{centred chromatic number} $\chicen(G)$ of $G$ is the minimum integer $c$ such that $G$ has a centred $c$-colouring.  We note that an easy argument shows that centred chromatic number and treedepth are equivalent, i.e., $\td(G)=\chicen(G)$ for every graph $G$. The \emph{linear chromatic number} of $G$ is the minimum integer $\ell$ such that $G$ has a linear $\ell$-colouring.  Since every path in $G$ is a connected subgraph of $G$, every centred colouring of $G$ is also a linear colouring of $G$, so $\chilin(G)\le\chicen(G)$.

@@

\begin{thm}[\citet{kun.obrien.ea:polynomial}]\label{kun_obrien_general}
  For any graph $G$, $\chicen(G)\le \chilin^{19}(G)\cdot\log^{O(1)}(\chilin(G))$.
\end{thm}

Their proof of \cref{kun_obrien_general} has three steps:
\begin{enumerate}
  \item A theorem of \citet{kawarabayashi.rossman:polynomial} shows that, if $\chicen(G)\ge k^{19}\log^{O(1)} k$ then $G$ contains a subcubic tree of treedepth $\Omega(k)$, or $\tw(G)\in\Omega(k^{18})$.  In the former case, an asymptotically optimal result of \citet{kun.obrien.ea:polynomial} on subcubic trees completes the proof, so we are left with the case where $\tw(G)\in\Omega(k^{18})$.
  \item The current-best version of the Excluded Grid Theorem due to \citet{chuzhoy:improved} shows that, if $\tw(G)\ge k^{18}$, then $G$ contains an $\Omega(k^2\times k^2)$ grid minor.\todo{Update reference}
  \item A technical lemma \cite[Lemma~5]{kun.obrien.ea:polynomial} shows that, if $G$ contains a $k^2\times k^2$ grid minor, then $\chilin(G)\in\Omega(k)$.\todo{Update reference}
\end{enumerate}

In the current work, we improve the technical lemma in Point~3 to show that a $k\times k$ grid minor is sufficient to guarantee that $\chilin(G)\in\Omega(k)$. This improves the exponent in \cref{kun_obrien_general} from $19$ to $10$.

This gives the following improvement to \cref{kun_obrien_general}
\begin{thm}\label{kun_obrien_general2}
  For any graph $G$, $\chicen(G)\in O(\chilin^{10}(G))$.
\end{thm}

In addition to the improvement to \cref{kun_obrien_general}, this results in considerably better bounds for some specific graph classes.  For example, the Excluded Grid Theorem for planar graphs asserts that any planar graph of treewidth $k$ contains a $\Omega(k)\times \Omega(k)$ grid minor.  Using this result instead of Point~2, above, implies that for any planar graph $G$, $\chicen(G)\in O(\chilin^2(G))$.

Finally, \citet{kun.obrien.ea:polynomial} boldly conjecture that, for any graph $G$, $\chicen(G)\le 2\chilin(G)$.  However, before the current work, the only class of graphs for which centred chromatic number was known to be bounded by a linear function of linear chromatic number was bounded degree trees.  The current result adds further evidence for this conjecture by establishing that, when $G$ is a $k\times k$ grid (or even a pseudogrid), then $\chicen(G)\in\Theta(\chilin(G))$.
% 
% These notes are an attempt to improve the bound in \cref{kun_obrien_general} in the general case as well as for some special classes of graphs. First we point out that some new results can already reduce the exponent from $190$ to $39$ and eliminate the polylog factor.  (These improvements are already described by \citet[Theorem~1.9]{czerwinski.nadara.ea:improved}.)
% 
% First, \citet{chuzhoy.tan:towards} improved the exponent in the Excluded Grid Theorem:  There exists a constant $c$ such that any graph $G$ with $\tw(G)\ge k^{18}\log^c k$ has a $k\times k$ grid minor.  This already improves the bound from $k^{190}$ to $k^{90}$.
% Next, a recent result of \citet[Theorem~1.4]{czerwinski.nadara.ea:improved} shows that if $b$ is the maximum treedepth of any tree contained in $G$ then $\td(G)\le b\cdot\tw(G)$ (and the proof is easy).  Therefore, if $\td(G)\ge k^{19}$ then at least one of the following cases applies:
% \begin{enumerate}
%     \item $\tw(G)\ge k^{18}$: In this case the proof follows Steps~2 and 3 as before, but using the grid minor result of \citet{chuzhoy.tan:towards} to show that $\chilin(G)\in\Omega(k)$.
%     \item $G$ contains a tree $T$ of treedepth $k$.  The same paper \cite[Lemma~1.6]{czerwinski.nadara.ea:improved} shows that any tree $T$ of treedepth $k$ contains a subcubic subtree of treedepth $\Omega(k)$.
%     A theorem of \citet{kun.obrien.ea:polynomial} shows that, for any subcubic tree $T$, $\chicen(T) \le (\log_2 3)\cdot\chilin(T)$.  Therefore, in this case $\chilin(G)\ge\chilin(T)\ge \chicen(T)/\log_2 3 = \Omega(k)$.
% \end{enumerate}

% As a next step, we would like to improve the third part of the argument to show that, if $G$ contains a $k\times k$ grid minor, then $\chilin(G)\in \Omega(k)$.  This, by itself, would reduce the exponent in \cref{kun_obrien_general2} from $39$ to $20$.  We observe that the first two parts of the argument are both ``heavy hammers'' and that, for some special graph classes, much lighter hammers suffice.  For example, if we consider only planar graphs, then it is well known that any planar graph $G$ of treewidth $k$ contains an $\Omega(k\times k)$ grid minor.  This already reduces the exponent in \cref{kun_obrien_general2} further (for planar graphs) to $3$ ($2$ if we can prove the result for grid minors).  If one could prove a similar result for the first ``heavy hammer'' then this would show that, for any planar graph $G$, $\chicen(G)\in O(\chilin(G))$.
%

@@

We will also make use of the following version of the Lovász Local Lemma:

\begin{lem}\label{weighted_lovasz}
  Let $E:=\{E_1,\ldots,E_n\}$ be a set of events in some probability space $(\Omega,\Pr)$.  For each $i\in\{1,\ldots,n\}$, let $\Gamma_i\subseteq E$ be such that the event $E_i$ is mutually independent of $E\setminus \Gamma_i$,\footnote{An event $A$ is mutually independent of a set $\{B_1,\ldots,B_r\}$ of events if, for any disjoint sets $I,J\subseteq\{1,\ldots,r\}$, $\Pr(A\cap\bigcap_{i\in I} B_i\cap\bigcap_{j\in J} \overline{B}_j=\Pr(A)\Pr(\bigcap_{i\in I} B_i\cap\bigcap_{j\in J} \overline{B}_j)$.} and let $w:E\to(0,1)$ be such that
  \[
      \Pr(E_i) \le w(E_i)\cdot\prod_{E_j\in\Gamma_i}(1-w(E_j))  \enspace .
  \]
  Then $\Pr(\overline{E_1}\cap\cdots\cap\overline{E_n}) > 0$.
\end{lem}

@@

In this work, we need to work with pseudogrids.  An \defin{$a\times b$-pseudogrid} is any graph that can be obtained from $G_{a\times b}$ in the following way:
\begin{compactitem}
  \item Replace each edge $vw$ of $G_{a\times b}$ with a path $\overline{P}_{vw}$ whose endpoints are $v$ and $w$.  (In other words, $\overline{P}_{vw}$ is a path obtained by subdividing the edge $vw$ zero or more times.)
  \item Replace each degree-$4$ vertex $(x,y)$ of $G_{a\times b}$ with a path $P_v$. If $P_v$ has only one vertex then the unique vertex of $P_v$ is adjacent to each of $(x\pm 1,y\pm 1)$ and this has effect.  Otherwise, $P_v$ has two endpoints $p$ and $q$, each of which is adjacent to two vertices among $(x\pm 1,y\pm 1)$.  It is useful to consider three possible cases: 
  \begin{compactenum}[(Q1)]
    \item \label{q_i} $p$ is adjacent to $\{(x-1,y), (x,y-1)\}$ and $q$ is adjacent to $(x+1,y),(x,y+1)$;
    \item \label{q_ii} $p$ is adjacent to $\{(x,y+1), (x-1,y)\}$ and $q$ is adjacent to $(x,y-1),(x+1,y)$; or
    \item \label{q_iii} $p$ is adjacent to $\{(x-1,y),(x+1,y)\}$ and $q$ is adjacent to $\{(x,y-1),(x,y+1)\}$.
  \end{compactenum}
\end{compactitem}

Let $G$ be an $a\times b$ pseudogrid.  For an edge $vw$ of $G_{a\times b}$, we let $P_{vw}:=\overline{P}_{vw}-\{v,w\}$ denote the (possibly empty) subpath containing the internal vertices of $\overline{P}_{vw}$.  For each vertex $v$ of $G_{a\times b}$ of degree less than $4$ we define $P_{v}$ to be the $1$-vertex path that contains only $v$.  In this way, $\mathcal{P}:=\{V(P_\mu):\mu\in \VE(G_{a\times b})\}$ is a partition of $V(G)$ into induced paths and the quotient graph $G/\mathcal{P}$ is isomorphic to $G_{a\times b}$.  We call $\mathcal{P}$ a \defin{grid-partition} of $G$.  
% and we denote it by $(P_\mu:\mu\in \VE(G_{a\times b}))$.
% This defines a mapping $\delta:V(G)\to V(G_{a\times b})\cup E(G_{a\times b})$ where $u\in V(P_{\delta(u)})$ for each $u\in v(G)$.

@@

Each row $R':=v_1,\ldots,v_a$ of $G_{a\times b}$ corresponds naturally to a path $R$ of $G$. The path $R$ contains $V(\overline{P}_{v_iv_{i+1}})$ for each $i\in\{1,\ldots,a-1\}$.  However, for $i\in\{2,\ldots,a-1\}$ $R$ may or may not contain $V(P_{v_i})$.  In particular, if $P_{v_i}$ was created using (Q\ref{q_iii}) then $R$ does not contain any internal vertices in $P_{v_i}$. Similarly, a column $C':=v_1,\ldots,v_b$ of $G_{a\times b}$ corresponds to a path that contains $V(\overline{P}_{v_jv_{j+1}})$ for each $j\in\{1,\ldots,b-1\}$. This correspondence allows us to talk about the rows and columns of $G$, which we will do next.

As part of our proof, we use the operation of \defin{deleting} a row (or column) $R$ of $G$, which works as follows:  We remove all edges of $R$ from $G$ and eliminate any isolated vertices.  If this produces vertices of degree $1$ (which happens when $R$ is the first or last row of $G$ or when $R=v_1,\ldots,v_a$ does not contains $P_{v_i}$ for some $i\in\{1,\ldots,r\}$) then we repeatedly remove degree-$1$ vertices until none remain.  If $G$ is an $a\times b$ pseudogrid and we delete some row $R$, then the resulting graph is an $a\times (b-1)$ pseudogrid.  Similarly, if we delete column $C$ of $G$, then the resulting graph is a $(a-1)\times b$ pseudogrid.

\subsection{Proof Outline}

If some graph contains a $k\times k$ grid minor then it contains a $k\times k$ pseudogrid as a subgraph.  Thus, our goal is to show that that $\chilin(G)\in \Omega(k)$ for any $k\times k$ pseudogrid.  We do this by showing that, any $\epsilon k$-colouring of any $k\times k$ pseudogrid $G$ contains an uncentred path $P$. We will do this in several steps (see \cref{outline}):

\begin{figure}
  \begin{center}
    \begin{tabular}{ccc}
      \includegraphics{figs/outline-2} &
      \includegraphics{figs/outline-3} &
      \includegraphics{figs/outline-4} \\
      $S_1$ & $S_1\cup S_2$ & $P$
    \end{tabular}
  \end{center}
  \caption{Constructing the sets $S_1$ and $S_2$ and the path $P$.}
  \label{outline}
\end{figure}
\begin{compactenum}

  \item We first delete rows and columns from $G$ so that each colour that appears in $G$ appears at least $C$ times, for some large constant $C$.  From this point on the goal is to construct $P$ so that it contains each colour at least twice.

  \item We do our best to find a set $S_1$ of vertices in $G$ that contains two vertices of each colour and that is `well-separated' in the sense that the corresponding set of vertices/edges in $G_{k\times k}$ have a minimum distance between them.  In \cref{outline} this minimum distance corresponds to the fact that the box drawn centered at each vertex in $S_1$ contains no other vertices of $S_1$. 
  % This makes it easy to find a single path in $G$ that contains each of these vertices.  
  Unfortunately, this process will fail for some subset of the colours that appear in $G$.

  \item For these failed colours, we use a variant of Hall's Theorem to match each failed colour $\alpha$ with two vertices in $S_1$ that each contain a vertex of colour $\alpha$ nearby.  In this matching, no vertex of $S_1$ is matched with more than one failed colour.  From this we identify a set $S_2$ that contains two vertices of each of the missing colours and such that no vertex of $S_2$ is close to any other vertex of $S_2$.

  \item We construct a path $P$ that contains each vertex in $S_1\cup S_2$.  This is possible (easy, even) because each vertex in $S_1$ is `close to' at most one vertex of $S_2$ and vice-versa.
\end{compactenum}

The most challenging aspect of this proof is the construction of $S_2$, which requires the use of the Local Lemma to ensure that no vertex chosen to take part in $S_2$ is close to any other vertex in $S_2$.  The difficulty is illustrated in \cref{outline} by the cluster of three points of $S_1$ in the top right corner.  These three points of $S_1$ are well-separated, but choosing one point from each of the three boxes to take part in $S_2$ could result in three points of $S_2$ being very close to each other.  The local lemma is used to show that, in situations like this, we can still obtain two vertices of $S_2$ from each missing colour class while avoiding taking two vertices of $S_2$ from these three boxes.

We now proceed with the various steps in the proof.

@@

For convenience, let $G_k:=G_{k\times k}$.
Let $G$ be a $k\times k$ pseudogrid with grid partition $\mathcal{P}:=(P_\mu:\mu\in\VE(G_{k}))$ and let $\varphi:V(G)\to\{1,\ldots,c\}$ be a vertex colouring of $G$.  The partition $\mathcal{P}$ associates each vertex of $G$ with an edge or vertex $\mu$ of $G_k$, so $\varphi$ associates a colour set with each object in $\VE(G_k)$, as follows.  For each $\mu\in\VE(G_{k})$, we let $\varphi_{\mathcal{P}}(\mu):=\{\varphi(v):v\in V(P_\mu)\}$.  For each colour $\alpha\in\{1,\ldots,c\}$, define $\varphi^{-1}(\alpha):=\{v\in V(G):\varphi(v)=\alpha\}$ and define $\varphi_\mathcal{P}^{-1}(\alpha):=\{\mu\in \VE(G_{k\times k}):\alpha\in\varphi_\mathcal{P}(\mu)\}$.  For any colour set $A\subseteq\{1,\ldots,c\}$ define $\varphi^{-1}(A):=\bigcup_{\alpha\in A}\varphi^{-1}(\alpha)$ and $\varphi_\mathcal{P}^{-1}(A):=\bigcup_{\alpha\in A}\varphi_{\mathcal{P}}^{-1}(\alpha)$.  (Throughout this section, we will use the idiom $\varphi(V(G))$ to denote the set of all colours used by $\varphi$ to colour the vertices of $G$.)

@@

\begin{proof}
  The proof is by induction on $|\varphi(V(G))|$, the number of colours used by the colouring $\varphi$.  If there exists no $A\subseteq \varphi(V(G))$ with $\varphi_{\mathcal{P}}^{-1}(A) < d|A|$ then taking $G':=G$ and $\mathcal{P}':=\mathcal{P}$ satisfies the requirements of the lemma. Otherwise, we will remove a set $R$ of rows and a set $C$ of columns from $G$ with $|R|=|C|\le d|A|$ to eliminate all vertices with colours in $A$, as follows:
  \begin{compactitem}
    \item For each $v:=(i,j)\in V(G_{k})$ with $A\cap\varphi_\mathcal{P}(v)\neq\emptyset$, we include row $j$ in $R$ and column $i$ in $C$.
    \item For each horizontal edge $vw\in E(G_{k\times k})$ with $A\cap\varphi_\mathcal{P}(vw)\neq\emptyset$, we include the row of $G$ that contains $P_{vw}$ in $R$.
    \item For each vertical edge $vw\in E(G_{k\times k})$ with $A\cap\varphi_\mathcal{P}(vw)\neq\emptyset$, we include the row of $G$ that contains $P_{vw}$ in $R$.\todo{Define horizontal and vertical edges of $G$.}
    \item Finally, we add arbitrary rows to $R$ or columns to $C$ to ensure that $|R|=|C|$.
  \end{compactitem}
  At this point $|R|=|C|\le d|A|$ and we remove all rows in $R$ and all columns in $C$ from $G$ to obtain a $k_0\times k_0$ pseudogrid $G_0$ with $k_0\ge k-d|A|$ and such that $\varphi(V(G_0))\cap A=\emptyset$.  In particular, $|\varphi(V(G_0))|\le |\varphi(V(G))|-|A|$.

  Now apply induction on $G_0$ to get a $k'\times k'$ pseudogrid with
  \[  k'\ge k_0-d|\varphi(V(G_0))|  \ge k-d|A|-d|\varphi(V(G_0))|\ge k - d|\varphi(V(G))|
  \]
  that satisfies the conditions of the lemma.
\end{proof}

Having each vertex or edge $\mu$ of $G_k$ associated with a set $\varphi_\mathcal{P}(\mu)$ of colours rather than a single colour will be problematic for what we want to to do next.  The following lemma allows us to choose one representative colour $\phi(\mu)$ from $\varphi_{\mathcal{P}}(\mu)$ for each $\mu\in \VE(G_k)$ while still ensuring that each colour appears frequently.

\begin{lem}\label{one_colour_per_object}
  Let $d>1$ be an integer, let $G$ be a $k\times k$ pseudogrid with grid-partition $\mathcal{P}$, and let $\varphi:V(G)\to\{1,\ldots,c\}$ be a vertex colouring of $G$ such that, for any $A\subseteq\varphi(V(G))$, $|\varphi_{\mathcal{P}'}^{-1}(A)| \ge d\cdot|A|$. Then there exists a colouring $\phi:\VE(G_{k})\to\{1,\ldots,c,\perp\}$ with the following properties:
  \begin{compactenum}[(i)]
    \item For each $\mu\in\VE(G_{k})$, $\phi(\mu)=\perp$ or $\phi(\mu)\in\varphi(\mu)$.
    \item For each $\alpha\in\varphi(V(G))$, $|\{\mu\in\VE(G_{k}):\phi(\mu)=\alpha\}|\ge d$.
  \end{compactenum}
\end{lem}

\begin{proof}
  Consider the bipartite graph $H$ with parts $X:=\{1,\ldots,c\}$ and $Y:=\VE(G_k)$ and edge set
  \[
    E(H) := \{ (\alpha,\mu)\in X\times Y: \alpha\in\varphi_\mathcal{P}(\mu) \}
  \]
  By \cref{d_hall}, $H$ contains a subgraph $M$ with $\deg_M(x)=d$ for each $\alpha\in X$ and $\deg_M(y)\le 1$ for each $\mu\in Y$.  For each edge $\alpha\mu$ in $M$, set $\phi(\mu):=\alpha$.  This defines $\phi(\mu)$ for any $\mu\in Y$ with $\deg_M(\mu)=1$.  For each $\mu\in Y$ with $\deg_M(\mu)=0$ set $\phi(\mu):=\perp$.
\end{proof}

@@

The \defin{$r$-box} centred at a vertex $v:=(i,j)$ of $G_{k}$ is defined as
\[
  B_r(v) := \{i-r,\ldots,i+r\}\times\{j-r,\ldots,j+r\} \cap V(G_{k}) \enspace .
\]
The \defin{$r$-box} centred at an edge $vw$ of $G_{k}$ is $B_r(vw):=B_r(v)\cup B_r(w)$.  
% Observe that, for any $\mu\in V(G_{k})\cup E(G_{a\times b})$,  $v\in B_r(u)$ if and only if $u\in B_r(v)$.  
Any $r$-box $B_r(\mu)$ defines an induced subgraph that we denote by $G_r(\mu):=G_{k}[B_r(\mu)]$.  We extend these definitions to vertices of an $k\times k$ pseudogrid $G$ with grid-partition $\mathcal{P}:=\{V(P_\mu):\mu\in\VE(G_k)\}$ as follows. For any $\mu\in V(G_k)$, define
\[
   \tilde{B}_r(\mu) := \bigcup_{\nu\in \VE(G_r(\mu))} V(P_\nu)
\]
and, for any $v\in V(G)$, let $\tilde{B}_r(v):=\tilde{B}_r(\mu_v)$
where $\mu_v$ denotes the unique element in $\VE(G_k)$ such that $v\in P_{\mu_v}$.\footnote{Technically, the notation for $\tilde{B}_r(v)$ should include the partition $\mathcal{P}$, but we omit this since there will never any ambiguity as to which partition is being used.}
% For any $v\in V(G)$, define $\tilde{G}_r(v):=G[\tilde{B}_r(v)]$.

@@

%   \begin{centre}

@@

%   \end{centre}
